---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all()
```

# wrswoR

<!-- badges: start -->
<!-- badges: end -->

The goal of wrswoR is to provide faster implementations of weighted random sampling without replacement in R.

## Installation

You can install the released version of wrswoR from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("wrswoR")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("krlmlr/wrswoR")
```
## Example

The functions in this package are a drop-in replacement to `sample.int(n, size, replace = FALSE, prob = prob)`.
With large `n`, `sample.int()` becomes too slow to be practical, unlike the functions in this package.

```{r example}
library(wrswoR)
set.seed(20200726)

sample_int_crank(20, 10, 1:20)
```
